Cryptanalysis of \MORUS

### 2018-05-15 morusAC_04_Mini_trails.tex
Mention src code as supportial material.
ePrint version: author names + acknowledgments + increased margins
added link to implementation
minor typos/clarifications

In this section, we describe how we build a trail for \MiniMORUS before computing its correlation and experimentally verify it.

@@

  \begin{subfigure}{.32\textwidth}
  \tikzsetnextfilename{minimorus_part_delta}
  \begin{tikzpicture}[xscale=0.65,yscale=1.5]%{{{
    \printstate
    \draw[trail, delta]
      (W-24) node[above] {$i$} -- (xnd4) (xnd4) -- (xor4) (xnd4) -- (and4)
      (xor4) -- (lll4) (xor4) -- (txor4) (txor4) -- (W42) node[below] {$i$}
      (lll4) -- (W44) node[below] {$i+b_4$}
      ;
  \end{tikzpicture}%}}}
  \caption*{$\delta^t_i$: weight 1} %$S^{j,4_i}, S^{j+1,2}_i, S^{j+1,4}_{i+13}$ ($w\!=\!1$)}
  \end{subfigure}
  \hfill
  \begin{subfigure}{.32\textwidth}
  \tikzsetnextfilename{minimorus_part_epsil}
  \begin{tikzpicture}[xscale=0.65,yscale=1.5]%{{{
    \printstate
    \draw[trail, epsil]
      (W-22) node[above] {$i$} -- (xnd2) (xnd2) -- (xor2) (xnd2) -- (and2)
      (xor2) -- (lll2) (xor2) -- (txor2) (txor2) -- (W40) node[below] {$i$}
      (lll2) -- (W42) node[below] {$i+b_2$}
      ;
  \end{tikzpicture}%}}}
  \caption*{$\varepsilon^t_i$: weight 1} %$S^{j,2}_i, S^{j+1,0}_i, S^{j+1,2}_{i+7}$ ($w\!=\!1$)}
  \end{subfigure}
  \hfill
  \begin{subfigure}{.32\textwidth}
    \small
    \begin{align*}
      \intertext{\MiniMORUS[640]:}
\alpha^t_i:&& C^t_i &\to S^{t+1}_{0,i+5} \\
\beta^t_i:&& C^t_i, S^t_{0,i} &\to S^t_{1,i} \\
\gamma^t_i:&& S^t_{1,i}, S^{t+1}_{1,i+31} &\to S^t_{4,i} \\
\delta^t_i:&& S^t_{4,i}, S^{t+1}_{4,i+13} &\to S^{t+1}_{2,i} \\
\varepsilon^t_i:&& S^t_{2,i}, S^{t+1}_{2,i+7} &\to S^{t+1}_{0,i}.
      \intertext{\MiniMORUS[1280]:}
\alpha^t_i:&& C^t_i &\to S^{t+1}_{0,i+13} \\
\beta^t_i:&& C^t_i, S^t_{0,i} &\to S^t_{1,i} \\
\gamma^t_i:&& S^t_{1,i}, S^{t+1}_{1,i+46} &\to S^t_{4,i} \\
\delta^t_i:&& S^t_{4,i}, S^{t+1}_{4,i+4} &\to S^{t+1}_{2,i} \\
\varepsilon^t_i:&& S^t_{2,i}, S^{t+1}_{2,i+38} &\to S^{t+1}_{0,i}.
    \end{align*}
    \caption*{\MiniMORUS instances}
  \end{subfigure}
  \caption{\MiniMORUS linear trail fragments.}
  \label{fig:minimorusfragments}
\end{figure}

@@

In the equation for \MiniMORUS[1280] from the previous section, each line on the right-hand side of the equality involves distinct $S^t_{i,j}$ terms, and each line has a weight of 1. By the Piling-Up Lemma, it follows that if we assume distinct $S^t_{i,j}$ terms to be uniform and independent, then the expression on the right-hand side has a weight of 16. Hence the linear combination of ciphertext bits on the left-hand side has a correlation of $2^{-16}$. The same holds for \MiniMORUS[640] (cf. \Cref{sec:traileq}).

@@

>>>>

In this section, we describe how we build a trail for \MiniMORUS, then compute its correlation and validate the correlation experimentally.

@@

  \begin{subfigure}{.32\textwidth}
  \centering
  \tikzsetnextfilename{minimorus_part_delta}
  \begin{tikzpicture}[xscale=0.65,yscale=1.5]%{{{
    \printstate
    \draw[trail, delta]
      (W-24) node[above] {$i$} -- (xnd4) (xnd4) -- (xor4) (xnd4) -- (and4)
      (xor4) -- (lll4) (xor4) -- (txor4) (txor4) -- (W42) node[below] {$i$}
      (lll4) -- (W44) node[below] {$i+b_4$}
      ;
  \end{tikzpicture}%}}}
  \caption*{$\delta^t_i$: weight 1} %$S^{j,4_i}, S^{j+1,2}_i, S^{j+1,4}_{i+13}$ ($w\!=\!1$)}
  \end{subfigure}
  \hfill
  \begin{subfigure}{.32\textwidth}
  \centering
  \tikzsetnextfilename{minimorus_part_epsil}
  \begin{tikzpicture}[xscale=0.65,yscale=1.5]%{{{
    \printstate
    \draw[trail, epsil]
      (W-22) node[above] {$i$} -- (xnd2) (xnd2) -- (xor2) (xnd2) -- (and2)
      (xor2) -- (lll2) (xor2) -- (txor2) (txor2) -- (W40) node[below] {$i$}
      (lll2) -- (W42) node[below] {$i+b_2$}
      ;
  \end{tikzpicture}%}}}
  \caption*{$\varepsilon^t_i$: weight 1} %$S^{j,2}_i, S^{j+1,0}_i, S^{j+1,2}_{i+7}$ ($w\!=\!1$)}
  \end{subfigure}
  \hfill
  \begin{subfigure}{.32\textwidth}
    \small
    \begin{align*}
      \intertext{\MiniMORUS[640]:}
\alpha^t_i:&& C^t_i &\to S^{t+1}_{0,i+5} \\
\beta^t_i:&& C^t_i, S^t_{0,i} &\to S^t_{1,i} \\
\gamma^t_i:&& S^t_{1,i}, S^{t+1}_{1,i+31} &\to S^t_{4,i} \\
\delta^t_i:&& S^t_{4,i}, S^{t+1}_{4,i+13} &\to S^{t+1}_{2,i} \\
\varepsilon^t_i:&& S^t_{2,i}, S^{t+1}_{2,i+7} &\to S^{t+1}_{0,i}.
      \intertext{\MiniMORUS[1280]:}
\alpha^t_i:&& C^t_i &\to S^{t+1}_{0,i+13} \\
\beta^t_i:&& C^t_i, S^t_{0,i} &\to S^t_{1,i} \\
\gamma^t_i:&& S^t_{1,i}, S^{t+1}_{1,i+46} &\to S^t_{4,i} \\
\delta^t_i:&& S^t_{4,i}, S^{t+1}_{4,i+4} &\to S^{t+1}_{2,i} \\
\varepsilon^t_i:&& S^t_{2,i}, S^{t+1}_{2,i+38} &\to S^{t+1}_{0,i}.
    \end{align*}
    \caption*{\MiniMORUS instances}
  \end{subfigure}
  \caption{\MiniMORUS linear trail fragments.}
  \label{fig:minimorusfragments}
\end{figure}

@@

In the equation for \MiniMORUS[1280] from the previous section, each line on the right-hand side of the equality involves distinct $S^t_{i,j}$ terms (in the sense that no two lines share a common term), and each line has a weight of 1. By the Piling-Up Lemma, it follows that if we assume distinct $S^t_{i,j}$ terms to be uniform and independent, then the expression on the right-hand side has a weight of 16. Hence the linear combination of ciphertext bits on the left-hand side has a correlation of $2^{-16}$. The same holds for \MiniMORUS[640] (cf. \Cref{sec:traileq}).

@@

\ifanonymous
The programs used to verify the bias experimentally have been
submitted as supporting material.
\else
The programs we used to verify the bias experimentally are available at:\\
\centerline{
\url{https://github.com/ildyria/MorusBias}
}
\fi

### 2018-05-15 morusAC_05_Full_trails.tex
typo
minor typos/clarifications

In the actual trail for (either version of) full \MORUS, this exact situation occurs when combining trail fragments $\beta^t_i$ and $\gamma^t_i$. Indeed, $\beta^t_i$ requires approximating the term $S^t_{2,i} \cdot S^t_{3,i}$, while $\gamma^t_i$ requires approximating the term $S^t_{2,i} \cdot S^t_{3,i-w}$ (cf. \Cref{fig:trailcollision}). While in \MiniMORUS, these terms cancel out, in the full \MORUS, when adding all symmetric copies of the trail, we end up with the sum:
\begin{align}
&S^t_{2,i} \cdot S^t_{3,i} \oplus S^t_{3,i} \cdot S^t_{2,i+w}
\oplus S^t_{2,i+w} \cdot S^t_{3,i+w} \oplus S^t_{3,i+w} \cdot S^t_{2,i+2w}\notag\\
\oplus\; &S^t_{2,i+2w} \cdot S^t_{3,i+2w} \oplus S^t_{3,i+2w} \cdot S^t_{2,i+3w}
\oplus S^t_{2,i+3w} \cdot S^t_{3,i+3w} \oplus S^t_{3,i+3w} \cdot S^t_{2,i}.\label{eq:8circle}
\end{align}
It may be observed that the products occurring in the equation above involve eight terms forming a ring. The weight of this expression can be computed by brute force, and is equal to $3$.

@@

In the case of \MORUS[640], collisions between rotation constants further complicate the analysis. Specifically, when using trail fragment $\beta^t_i$, the term $S^t_{2,i} \cdot S^t_{3,i}$ occurs. As explained previously, a partial collision with the term $S^t_{2,i} \cdot S^t_{3,i-w}$ from trail fragment $\gamma^t_i$ results in \Cref{eq:8circle}. However trail fragment $\alpha^t_{i+d}$ is once used in the course of the full trail with an offset of $d = b_1+b_4-b_0-b_2$ (relative to $\gamma^t_i$), which in the case \MORUS[640] is equal to $31+13-5-7 = 0 \;\text{mod}\; 32$. This creates another term $S^t_{2,i} \cdot S^t_{3,i}$, which ultimately destroys one of the four occurrences of \Cref{eq:8circle}. Therefore, when computing the full trail equation on \MORUS[640], we get that the weight of the trail is 73 (cf. \Cref{sec:traileq}).

@@

To see this, recall that plaintext bits contribute to the encryption process in two ways (cf. \Cref{subsec/Spec}):
\begin{enumerate}
\item They are added to some bits derived from the state to form the ciphertext.\label{item:1}
\item During each encryption step, the \StateUpdate function adds a plaintext block to every register except $S_0$.\label{item:2}
\end{enumerate}

>>>>

In the actual trail for (either version of) full \MORUS, this exact situation occurs when combining trail fragments $\beta^t_i$ and $\gamma^t_i$. Indeed, $\beta^t_i$ requires approximating the term $S^t_{2,i} \cdot S^t_{3,i}$, while $\gamma^t_i$ requires approximating the term $S^t_{2,i} \cdot S^t_{3,i-w}$ (cf. \Cref{fig:trailcollision}). While in \MiniMORUS, these terms cancel out, in the full \MORUS, when adding all symmetric copies of the trail, we end up with the sum:
{
\allowdisplaybreaks[0]
\begin{align}
&S^t_{2,i} \cdot S^t_{3,i} \oplus S^t_{3,i} \cdot S^t_{2,i+w}
\oplus S^t_{2,i+w} \cdot S^t_{3,i+w} \oplus S^t_{3,i+w} \cdot S^t_{2,i+2w}\notag\\
\oplus\; &S^t_{2,i+2w} \cdot S^t_{3,i+2w} \oplus S^t_{3,i+2w} \cdot S^t_{2,i+3w}
\oplus S^t_{2,i+3w} \cdot S^t_{3,i+3w} \oplus S^t_{3,i+3w} \cdot S^t_{2,i}.\label{eq:8circle}
\end{align}
}
It may be observed that the products occurring in the equation above involve eight terms forming a ring. The weight of this expression can be computed by brute force, and is equal to $3$.

@@

In the case of \MORUS[640], collisions between rotation constants further complicate the analysis. Specifically, when using trail fragment $\beta^t_i$, the term $S^t_{2,i} \cdot S^t_{3,i}$ occurs. As explained previously, a partial collision with the term $S^t_{2,i} \cdot S^t_{3,i-w}$ from trail fragment $\gamma^t_i$ results in \Cref{eq:8circle}. However trail fragment $\alpha^t_{i+d}$ is once used in the course of the full trail with an offset of $d = b_1+b_4-b_0-b_2$ (relative to $\gamma^t_i$), which in the case of \MORUS[640] is equal to $31+13-5-7 = 0 \;\text{mod}\; 32$. This creates another term $S^t_{2,i} \cdot S^t_{3,i}$, which ultimately destroys one of the four occurrences of \Cref{eq:8circle}. Therefore, when computing the full trail equation on \MORUS[640], we get that the weight of the trail is 73 (cf. \Cref{sec:traileq}).

@@

To see this, recall that plaintext bits contribute to the encryption process in two ways (cf. \Cref{subsec/Spec}):
\begin{enumerate}
\item They are added to some bits derived from the state to form the ciphertext.\label{item:1}
\item During each encryption step, the \StateUpdate{} function adds a plaintext block to every register except $S_0$.\label{item:2}
\end{enumerate}
